### 2020-06-16 report/sections/background/sessiontypes.tex
Report: complete ch2.1

% need LTS semantics here too

>>>>

Whilst binary session types provide communication guarantees 
between exactly 2 participants, 
distributed systems generally involve more parties in practice. 
This is equally relevant in interactive web applications, 
as motivated by the \textit{Battleships} game example 
in \cite{PureScript2019} where the server coordinates 
interactions between two players.

Whilst there is a natural syntactical extension to 
our session calculus for 
describing multiparty sessions\footnote{
We also adopt the shorthand 
$\mathcal{M} ::= \prod^n_{i=1} \ptprocess{p_i}{P_i}$ 
as used in the literature.} as

\[
\begin{array}{rl}
\mathcal{M} ::= & \ptprocess{p_1}{P_1} ~\mid~
\ptprocess{p_2}{P_1} ~\mid~\dots~\mid
\ptprocess{p_n}{P_n}
\end{array}
\]

the same cannot be said for the binary session typing discipline,
particularly with respect to duality.
The same notion of duality does not extend to the 
decomposition of multiparty interactions into 
multiple binary sessions: 
\cite{C406Lecture} and \cite{MPST} both present 
counterexamples of well-typed binary sessions that, 
when composed to represent a multiparty session, 
results in communication errors thus 
violating guarantees of well-typed sessions.

Honda et al. presents {multiparty session types} 
in \cite{MPAST} to extend the 
binary session typing discipline for 
sessions involving more than 2 participants, 
whilst redefining the notion of compatibility 
in this multiparty context. 
\textit{Multiparty session types} (MPST)
are defined in terms a \textit{global type},
which provides a bird's eye view of the communication protocol
describing the interactions between pairs of participants. 
\cref{fig:mpst} defines the syntax of global types
inspired by \cite{MPST};
we omit message payload types for brevity.

% Introduce global type, projection
\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

G ::= & & \text{Global Types} \\
     & \tend & \text{Termination} \\
\mid & \gcomm{p}{q}{l_i: G_i}{i \in I} & \text{Message Exchange} \\
\mid & \trec{G} & \text{Recursive Type} \\
\mid & \trecvar & \text{Type Variable} \\
\end{array}
\]
\singlespacing
\caption{Syntax of Global Types}
\label{fig:mpst}
\end{figure}

To check the conformance of a participant's process 
against the protocol specification 
described by the global type, 
we \textit{project} the global type onto the participant
to obtain a session type that only preserves the 
interactions described by the global type that 
pertain to the participant. 
Projection is defined by the $\upharpoonright$ operator, 
more commonly seen in literature in its infix form 
as $\proj{G}{p}$ describing the projection of global type 
$G$ for participant $\role{p}$. 
Intuitively, the projected local type of a participant 
describes the protocol from the viewpoint of the participant.

More formally, projection can be interpreted as a 
\textit{partial function} 
${\upharpoonright} :: G \times \mrole{p} \rightharpoonup S$,
as the projection for a participant may be undefined for 
an ill-formed global type; 
\cite{C406Lecture} presents examples of where this is the case, 
and \cite{MPST} presents the formal definition of projection.
We focus on the definition of projection for the 
message exchange global type given below:

\[
\proj{\gcomm{p}{q}{l_i: G_i}{i \in I}}{r} =
\begin{cases}
\tsel{q}{l_i: \proj{G_i}{p}}{i \in I} 
	& \text{if } \mrole{r} = \mrole{p} \\
\tbra{p}{l_i: \proj{G_i}{q}}{i \in I} 
	& \text{if } \mrole{r} = \mrole{q} \\
\underset{i \in I}{\MERGEOP}~\proj{G_i}{r}
	& \text{otherwise} \\
\end{cases}
\]

The third case describes that,
for a participant not involved in the communication,
the projections of the continuations $\proj{G_i}{r}$
do not necessarily have to be the same, but they
need to be ``compatible''. 
The $\mergeop$ syntax denotes the \textit{merging operator}.
It is a partial function on local types allows us to construct
a projection that is compatible with the projections on
the continuations.
We use the \textit{inductive} definition presented by
Scalas and Yoshida in \cite{LessIsMore}, shown in \cref{fig:merge}.

\begin{figure}[!h]
\doublespacing
\[
\begin{array}{rlr}

\tmerge{\tend}{\tend} &= \tend 
	& \rulename{Merge-End} \\
\tmerge{\trecvar}{\trecvar} &= \trecvar 
	& \rulename{Merge-RecVar} \\
\tmerge{\trec{T}}{\trec{T'}} &= \trec{\tmerge{T}{T'}} 
	& \rulename{Merge-Rec} \\
\tmerge{\tsel{p}{l_i: T_i}{i \in I}}{\tsel{p}{l_i: T_i}{i \in I}} 
	&= \tsel{p}{l_i: T_i}{i \in I}
	& \rulename{Merge-Sel} \\
\tmerge{\tbra{p}{l_i: T_i}{i \in I}}{\tbra{p}{l_j: T'_j}{j \in J}} 
	&= \tbra{p}{l_k: T''_k}{k \in I \cup J}
	& \rulename{Merge-Bra} \\

	& 
	\text{where } T''_k =
	\begin{cases}
	T_k & \text{if } k \in (I \setminus J) \\
	T'_k & \text{if } k \in (J \setminus I) \\
	\tmerge{T_k}{T'_k} & \text{otherwise} \\
	\end{cases}
	& \\

\end{array}
\]
\singlespacing
\captionof{figure}{Definition of Merging Operator}
\label{fig:merge}
\end{figure}

The notion of compatibility in multiparty session types 
is still captured by \rulename{MTy}, 
but adapted to consider the local projections for 
all participants as supposed to dual types in the binary case.

\begin{prooftree}
\AxiomC{$
\forall i \in I. ~ \left( \cdot \vdash P_i : \proj{G}{p_i} \right)
$}
\AxiomC{$
\pt{G} \subseteq \left\{ \mrole{p_i} \mid i \in I \right\}
$}
\RightLabel{\rulename{MTy}}
\BinaryInfC{$
\displaystyle \vdash \prod^{}_{i \in I} \ptprocess{p_i}{P_i} ~: G
$}
\end{prooftree}

For a multiparty session 
$\mathcal{M} = \prod_{i \in I} \ptprocess{p_i}{P_i}$ 
to be well-typed by a global type $G$:

\begin{enumerate}
\item 
All participant processes $\mrole{p_i}{P_i}$ are well-typed 
with respect to their corresponding well-defined projection 
$\proj{G}{p_i}$, and 
\item 
$G$ does not describe interactions with participants 
not defined in $\mathcal{M}$.
$\pt{G}$ denotes the set of participants in the global type $G$,
defined inductively on $G$.

\begin{align*}
\pt{\tend} &= \emptyset \\
\pt{\trecvar} &= \emptyset \\
\pt{\trec{G}} &= \pt{G} \\
\pt{\gcomm{p}{q}{l_i: G_i}{i \in I}} &= 
	\{ \mrole{p}, \mrole{q} \} \cup \underset{i \in I}{\bigcup}~\pt{G_i} \\
\end{align*}

\end{enumerate}

Well-typed multiparty session enjoys the following 
guarantees as outlined in \cite{GentleMPST}:

\begin{itemize}
\item \textbf{Communication safety}: The types of sent and expected messages will always match.
\item \textbf{Protocol fidelity}: The exchange of messages between processes will abide by the global protocol.
\item \textbf{Progress}: Messages sent by a process will be eventually received, and a process waiting for a message will eventually receive one; this also means there will not be any sent but unreceived messages.
\end{itemize}

This motivates an elegant, decentralised solution for checking protocol conformance in practice: once the global type for the protocol is defined, local processes can verify their implementation against their corresponding projection in isolation, independent of each other. 
We illustrate this in \cref{fig:mpstworkflow}.

\begin{figure}[!ht]
\centering
\includegraphics[width=.75\textwidth]{MpstFramework}
\caption{Type Checking with Multiparty Session Types}
\label{fig:mpstworkflow}
\end{figure}

### 2020-06-16 report/sections/general/p2p/challenges.tex
Report: complete ch2.1

\begin{example}[Naive definition of routing in global types]
Consider the global type
\[
A \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
By existing LTS semantics over
global types (\cref{subsubsection:lts}), then $SB!\tmsg{M2}$
is a prefix of a valid execution trace.
If we define our routing construct simply by replacing every single
$A \to B: \tmsg{M}$ interaction with $A \to S: \tmsg{M}.
S \to B: \tmsg{M}$, we transform the global type into
\[
A \to S: \tmsg{M1}. S \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
our naive definition of routing has over-serialised the protocol
since $SB!\tmsg{M2}$ can no longer occur first because \trole{S}
is forced to send \tmsg{M1} first, which it cannot do so until receiving
from \trole{A}.
This implies that we need to handle routing interactions differently
from ``normal'' send/receive actions.
\end{example}

>>>>

\begin{example}[Naive definition of routing in global types]
Consider the global type
\[
A \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
By existing LTS semantics over
global types \cite{characterisation}, 
then $\aout{S}{B}{\tmsg{M2}}$
is a prefix of a valid execution trace.
If we define our routing construct simply by replacing every single
$A \to B: \tmsg{M}$ interaction with $A \to S: \tmsg{M}.
S \to B: \tmsg{M}$, we transform the global type into
\[
A \to S: \tmsg{M1}. S \to B: \tmsg{M1}. S \to B: \tmsg{M2}. \texttt{end}
\]
our naive definition of routing has over-serialised the protocol
since $\aout{S}{B}{\tmsg{M2}}$ can no longer occur first because \trole{S}
is forced to send \tmsg{M1} first, which it cannot do so until receiving
from \trole{A}.
This implies that we need to handle routing interactions differently
from ``normal'' send/receive actions.
\end{example}
